%	$Id: conclusions.tex,v 1.22 2012/08/20 10:04:22 pooka Exp $
%

\section{Conclusions}
\label{chap:conclusions}

We set out with the goals of improving characteristics of monolithic
kernel operating systems, namely security, code reusability and
testing and development characteristics.  Our view is that a
codebase's real value lies not in the fact that it exists, but that
it has been proven and hardened ``out there''.
Any approach which required us to start building the codebase from
scratch for an alternate kernel architecture was not a viable
solution for us.  In contrast, our work was driven by the needs of
the existing codebase, and how to gain maximal use out of it.

We claimed that a properly architected and maintained monolithic
kernel provides a flexible codebase.  We defined an \textit{anykernel}
to be an organization of kernel code which allows the kernel's
\textit{unmodified} drivers to be run in various configurations
such as application libraries and network servers, and also in the
original monolithic kernel.  We showed by implementation that the
NetBSD monolithic kernel could be turned into an anykernel with
relatively simple modifications.

An anykernel can be instantiated into units which virtualize the
minimum support functionality for kernel drivers; support which can be
used directly from the host is used directly from the host without a
layer of indirection.  The virtualized kernel driver instances
are called \textit{rump kernels} since they retain only a part of the
original features.  Our implementation hosts rump kernels in a process
on a POSIX host.  We discuss other possible hosts at the end of this
chapter alongside future work.

The development effort for turning a monolithic kernel into an
anykernel is insignificant next to the total effort a real world
capable monolithic kernel has received.  The anykernel architecture by
itself is not a solution to the problems; instead, it is merely a concept
which enables solving the problems while still retaining the ability to
run the kernel in the original monolithic configuration.

At runtime, a rump kernel can assume the role of an application library
or the role of a server.  Programs requesting services from rump kernels are
called rump kernel clients or simply clients.  We defined three client
types and implemented support for them.  Each client type maps to the
best alternative for solving one of our motivating problems.

\begin{enumerate}
\item	Local: the rump kernel is used in a library capacity.
	Requests are issues as local function calls.

\item   Microkernel: the host routes client requests from regular
	processes to drivers running in isolated servers.

\item   Remote: the client and rump kernel are running in
	different containers (processes) with the client deciding
	which services to request from the rump kernel.  The kernel
	and client can exist either on the same host or on
	different hosts and communicate over the Internet.
\end{enumerate}

Local clients allow using kernel drivers as libraries for applications,
where everything runs in a single process.  This execution model is not
only fast, but also convenient, since to an outside observer it looks
just like an application, \ie it is easy to start, debug and stop one.

Microkernel servers allow running kernel code of questionable stability in
a separate address space with total application transparency.  However,
they require both host support for the callback protocol (\eg VFS for
file systems) and superuser privileges to configure the service (in the
case of file systems: mount it).

Remote client system call routing is configured for each client
instance separately.  However, doing so does not require privileges,
and this approach is the best choice for virtual kernels to be used in
testing and development.

The key performance characteristics of rump kernels are:

\begin{itemize}
\item	Bootstrap time until application-readiness is in the order of
	10ms.  On the hardware we used it was generally under 10ms.

\item	Memory overhead depends on the components
	included in the rump kernel, and is typically from 700kB to 1.5MB.

\item   System call overhead for a local rump kernel is 53\% of a
	native system call for uniprocessor configurations and 44\%
	for a dual CPU configuration (\ie a rump kernel performs
	a null system call twice as fast), and less than 1.5\% of
	that of User-Mode Linux (the rump kernel syscall mechanism
	is over 67 times as fast).
\end{itemize}

Our case study for improving security was turning a file system
mount using an in-kernel driver to one using a rump kernel server.
From the user perspective, nothing changes.  From the system
administrator perspective, the only difference is one extra flag
which is required at mount time (\texttt{-o rump}).  Using isolated
servers prevents corrupt and malicious file systems from damaging
the host kernel.  Due to the anykernel architecture, the ability
to use the same file system drivers in kernel mode is still possible.

Our application case study involved a redo of the \textit{makefs}
application, which creates a file system image out of a directory
tree without relying on in-kernel drivers.  Due to the new ability of
being able to reuse the kernel file system drivers directly in
applications, our reimplementation was done in under 6\% of the
time the original implementation required.  In other words, it
required days instead of weeks to implement.  For another example
of using a rump kernel in an application, see
Appendix~\ref{sect:cgdencrypt} where using the in-kernel crypto
driver for creating an encrypted disk image is explained
using binaries available on an out-of-the-box NetBSD installation.

For testing and development we added close to 1,000 test cases using rump
kernels as the test backends.  The test suite is run against daily NetBSD
changes and has been able to find real regressions, including kernel
panics which would have caused the test suite to fail if testing was done
against the test host kernel.  Testing against rump kernels means that a
kernel panic is inconsequential to the test run and allows the test run to
complete in constant time irrespective of how many tests cause a kernel
crash.  Additionally, when problems are discovered, the test program
that was used to discover the problem can directly be used to debug
the relevant kernel code without the need to set up a separate kernel
debugging environment.

\subsection{Future Directions and Challenges}
\label{chap:future}

The hosting of NetBSD-based rump kernels on other POSIX-style operating
systems such as Linux was analyzed.  Running a limited set of applications
was possible, but interfacing between the host and rump namespaces is not
yet supported on a general level.  For example, the file status structure
is called \texttt{struct~stat} in the client and rump kernel namespaces,
but the binary layouts may not match, since the client uses the host
representation and the rump kernel uses the NetBSD representation.
If a reference to a mismatching data structure is passed from the client
to the rump kernel, the rump kernel will access incorrect fields.
Data passed over the namespace boundary need to be translated.  NetBSD
readily contains system call parameter translation code for a number of
operating systems such as Linux and FreeBSD under \texttt{sys/compat}.
We wish to investigate using this already existing translation code
where possible to enlarge the set of supported client ABIs.

Another approach to widening host platform support is to port rump
kernels to a completely new type of non-POSIX host.  Fundamentally,
the host must provide the rump kernel a single memory address space and
thread scheduling.  Therefore, existing microkernel interfaces such as L4
and Minix make interesting cases.  Again, translation may be needed when
passing requests from the clients to NetBSD-based rump kernels.  However,
since interface use is likely to be limited to a small subset, such as
the VFS/vnode interfaces, translation is possible with a small amount
of work even if support is not readily provided under \texttt{sys/compat}.

Multiserver microkernel systems may want to further partition a rump
kernel.  We investigated this partitioning with the \textit{sockin}
facility.  Instead of requiring a full TCP/IP stack in every rump kernel
which accesses the network, the sockin facility enables a rump kernel
to communicate its intentions to a remote server which does TCP/IP.
In our case, the remote server was the host kernel.

The lowest possible target for the rump kernel hypercall layer is firmware
and hardware.  This adaption would allow the use of anykernel drivers both
in bootloaders and lightweight appliances.  A typical firmware does not
provide a thread scheduler, and this lack would either mandate limited
driver support, \ie running only drivers which do not create or rely
on kernel threads, or the addition of a simple thread scheduler in the
rump kernel hypervisor.  If there is no need to run multiple isolated
rump kernels, virtual memory support is not necessary.

An anykernel architecture can be seen as a gateway from
current all-purpose operating systems to more specialized operating
systems running on ASICs.  Anykernels enable the
device manufacturer to provide a compact hypervisor and select only
the critical drivers from the original OS for their purposes.  The
unique advantage is that drivers which have been used and proven
in general-purpose systems, \eg the TCP/IP stack, may be included
without modification as standalone drivers in embedded products.
